---
title: "Stats 330 Assignment 3"
author: "Hasnain Cheena, 190411106, hche737"
date: "18/04/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
library(mgcv)
library(MASS)
library(statmod)
library(MuMIn)
```

```{r Data Import and Setup, echo=FALSE}
nyBridges.df = read.csv("NYbridges.csv")
```


## Question 1

```{r Question 1 - Communication}

#convert degrees fahrenheit to celsius
HighC = (nyBridges.df$High.Temp - 32) * 5/9
LowC = (nyBridges.df$Low.Temp - 32) * 5/9

#convert rain in inches to mm
#as a result of this conversion trace amounts (denoted with a T) are converted to NaN
Rainmm = as.numeric(as.character(nyBridges.df$Precipitation)) * 25.4

#relevel day variable
day.relevel = relevel(nyBridges.df$Day, "Monday")
```
*Day Factor Relevel*  
Monday was selected to be the baseline level as in New Zealand we consider Monday to be the first day of the week. Further by using Monday we can easily assess if there is a difference in bicycle travel between the first day of week and the rest of the week. 

*Trace Characteristics*  
Trace characterises an amount of percipitation that is greater than 0 but too small to be measured reliably. The "T" used in the dataset denotes trace amounts of  percipitation. As I am not a meterologist and therefore cannot confirm whether trace amounts of perciptation are equivalent to 0 and so I have chosen to recode trace amounts as NaN.  

Reference for definition of trace: https://www.wmo.int/pages/prog/www/IMOP/publications/CIMO-Guide/Prelim_2018_ed/8_I_6_en_MR_clean.pdf

## Question 2

```{r Question 2 - Create new dataframe, echo=FALSE}
nyBridges.Q2.df = data.frame(Brooklyn=nyBridges.df$Brooklyn, 
                            Rainmm = Rainmm, 
                            HighC = HighC, 
                            LowC = LowC, 
                            Day = day.relevel)
```


```{r Question 2 - Brooklyn Bridge}
pairs20x(nyBridges.Q2.df, main="Pairs Plot")
```
(a) The statistician did not include both high temperature ($HighC$) and low temperature ($LowC$). This is because as you can see from the pairs plot that the high temperature variable and low temperature variable are highly correlated. Thus, both HighC and LowC explain the same pattern in the response variable and therefore the statistician only included one of them. HighC in specific was choosen HighC was choosen as HighC has a stronger correlation to the response than LowC.  

(b) Firstly, the log transformation was used on Rainmm as the variable is very right skewed (shown on pairs plot histogram). However using the natural log transformation will not work as Rainmm contains the value 0. Therefore log1p was used as it log1p computes the computes the natural logarithm of the given value plus 1 accounting for Rainmm values of 0. 


(c) A quadratic effect for temperature was included in the models because the relatioship between temperature ($HighC$) and the response showed curvature indicating a non-linear relationship. Therefore a quadratic term for temperature was added to account for the non-linearity aspect of the relationship.

## Question 3
```{r Question 3, echo=FALSE}

NYBridges.df = data.frame(Brooklyn=nyBridges.df$Brooklyn, 
                            Rainmm = Rainmm, 
                            HighC = HighC, 
                            LowC = LowC, 
                            Day = day.relevel)


```

### Model A Exploration

```{r Question 3 - Model A Exploration}

model.lin.a<-lm(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),data=NYBridges.df)

layout(matrix(c(1,1,1,1,1,1,2,2,2,2,2,2), nrow = 4, ncol = 3, byrow = TRUE))
#raw residual
plot(predict(model.lin.a),residuals(model.lin.a), ylab="Raw Residuals",xlab="Fitted values")
abline(h=0, lty='dashed')
#pearson residual 
plot(predict(model.lin.a),residuals(model.lin.a, type="pearson"), ylab="Pearson Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')
normcheck(model.lin.a, shapiro.wilk = TRUE)

#assess goodness of fit
deviance(model.lin.a)
```

The raw residuals and pearson residuals do not show a patternless band around 0. The pearson residuals clearly display non-constant variance and the majority of the residuals are not between -2 and 2. Furthermore, the histogram of residuals shows the residuals are very right skewed.
The residual sum of squares is 46551228. This value was not useful in making a decision on the model as we had no scale to compare to inorder to make the decision that the model was appropriate. 

All the evidence above coupled with the fact that the response variable is a count lead to the conclusion that the linear model is not appropriate to capture the relationship correctly between the bicycle activity on the Brooklyn bridge and explanatory variables. 

### Model B Exploration

```{r Question 3 - Model B Exploration}

model.pois.b<-glm(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),family=poisson,data=NYBridges.df)

layout(matrix(c(1,1,1,1,1,1,2,2,2,2,2,2), nrow = 4, ncol = 3, byrow = TRUE))
#pearson residual 
plot(predict(model.pois.b),residuals(model.pois.b, type="pearson"),ylab="Pearson Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')
#deviance 
plot(predict(model.pois.b),residuals(model.pois.b, type="deviance"),ylab="Deviance Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')

#assess goodness of fit
1-pchisq(model.pois.b$deviance, model.pois.b$df.residual)
```
The pearson and deviance residual plots do not have a patternless band around 0. Therefore, the residual plots show non-constant variance. Furthermore, a majority of the residuals are not within -2 and 2 indcating the model is not appropriate.
Moreover, we have very strong evidence (p-value = 0) that the model is not a good fit. 
Taking the above facts into consideration this model is also not appropriate to capture the relationship correctly between the bicycle activity on the Brooklyn bridge and explanatory variables. 

### Model C Exploration

```{r Question 3 - Model C Exploration}

model.qpois.c<-glm(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),family=quasipoisson,data=NYBridges.df)

layout(matrix(c(1,1,1,1,1,1,2,2,2,2,2,2), nrow = 4, ncol = 3, byrow = TRUE))
#pearson residual 
plot(predict(model.qpois.c),residuals(model.qpois.c, type="pearson"),ylab="Pearson Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')
#deviance 
plot(predict(model.qpois.c),residuals(model.qpois.c, type="deviance"),ylab="Deviance Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')
```
The Pearson and deviance residual plots are still showing non-constant variance as they are not a patternless band around 0.  This suggests that the quasi-Poisson model is not suitable to model the relationship between the bicycle activity on the Brooklyn bridge and explanatory variables. 

### Model D Exploration

```{r Question 3 - Model D Exploration}

model.nb.d<-glm.nb(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),data=NYBridges.df)

layout(matrix(c(1,1,1,1,1,1,2,2,2,2,2,2), nrow = 4, ncol = 3, byrow = TRUE))
#pearson residual 
plot(predict(model.nb.d),residuals(model.nb.d, type="pearson"),ylab="Pearson Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')
#deviance 
plot(predict(model.nb.d),residuals(model.nb.d, type="deviance"),ylab="Deviance Residuals", xlab="Fitted values")
abline(h=0, lty='dashed')

#summary
summary(model.nb.d)

#assess goodness of fit
1-pchisq(model.nb.d$deviance, model.nb.d$df.residual)
```
The Pearson and deviance residuals are scattered in a reasonably patternless band around 0. Further, a majority of the residuals are between -2 and 2. Therefore, the residuals do not show a pattern in the mean and variance. 
Further, the deviance does not provide evidence (p-value \approx 0.24) for a lack of fit. 
The negative binomial model is appropriate to model the relationship between the bicycle activity on the Brooklyn bridge and explanatory variables. 

## Question 4

```{r Question 4 - Part 1}
#weekend vs weekday
#relevel with baseline as sunday
NYBridges.df$Day <- relevel(NYBridges.df$Day, "Saturday")
model.nb.d<-glm.nb(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),data=NYBridges.df)
summary(model.nb.d)
#relevel with baseline as saturday
NYBridges.df$Day <- relevel(NYBridges.df$Day, "Sunday")
model.nb.d<-glm.nb(Brooklyn~log1p(Rainmm)+Day+HighC+I(HighC^2),data=NYBridges.df)
summary(model.nb.d)

100*(exp(confint(model.nb.d))-1)
```
*Brief Model Interpretation*
There is no evidence of a difference between the number of cyclists in Brooklyn on Saturday compared to all the weekdays.

There is evidence of the a difference between the number of cyclists in Brooklyn on Sunday compared to Tuesday, Wednesday and Thursday. However there is no evidence of a difference between the number of cyclists in Brooklyn on Sunday compared to Monday and Friday. 
We estimate that, for the same percipitation amount and temperature, the expected number of cyclists in Brooklyn on Tuesday is 8.5% to 41.5% higher than Sunday. 
We estimate that, for the same percipitation amount and temperature, the expected number of cyclists in Brooklyn on Wednesday is 5.7% to 37.9% higher than Sunday. 
We estimate that, for the same percipitation amount and temperature, the expected number of cyclists in Brooklyn on Thursday is 1.3% to 32.2% higher than Sunday. 

*Which of the four models do you think is best?*
From the evidence presented above the negative binomial model ($model.nb.d$) is the best. Relative to the other models the Pearson and deviance residual plots of the negative binomial model show reasonably constant variance and no pattern in the mean. Further most of the residuals are within -2 and 2. Thus the modelling assumptions are met unlike the other models. Further, for the negative binomial model the deviance does not provide evidence (p-value \approx 0.24) for a lack of fit. In constrast all the other models specified provide evidence against the hypothesis that the model is a good fit. 

```{r Question 4 - Part 2}
AIC(model.lin.a,model.pois.b,model.nb.d)
AICc(model.lin.a,model.pois.b,model.nb.d)
BIC(model.lin.a,model.pois.b,model.nb.d)
```

*Do you think the model fits well?*
I believe the negative binomial model ($model.nb.d$) is a good fit but it does not fit the data well. This is because of the AIC, AICc and BIC scores. The AICc, AIC and BIC scores show that the linear model ($model.lin.a$) is considerably better supported than by the data than the the negative binomial model ($model.nb.d$). However, we know the linear model is not appropriate and does not fit the data well. Thus since it is least-worst out of the candidate models, all the candidate models must not fit the data well. 

*Other variables*
Additional meterological variables I believe this dataset could benefit from are:
* Cloud cover: Days with high cloud cover may indicate chance of rain and thus de-incentize cyclists. 
* Average Wind speed - Days with higher wind speed makes bicycle riding harder increasing the possibilty of less cyclists on the Brooklyn bridge. 
* Hours of sunshine - Hours of daily sunshine could give an indication on the season. Less daily sunshine could mean winter and thus less cyclists on the Brooklyn bridge due to the cold conditions. 
* Relative Humidity - Quite humid days are uncomfortable may deter cyclists. 


