---
title: "Stats  330 Assignment 1"
author: "Hasnain Cheena"
date: "16/03/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
```

## Data Cleaning

```{r Data Cleaning}
#read in data

heartHealth.df = read.csv("hearthealth.csv")
#make a copy for this question
heartHealth.cleaned.df = heartHealth.df[c("height", "weight", "sbp", "dbp")]

#initial pairs plot
pairs20x(heartHealth.cleaned.df, main="Initial Pairs Plot") 

#adjustment 1
heartHealth.cleaned.df[heartHealth.cleaned.df["height"] == 15.1, "height"] = 1.51
pairs20x(heartHealth.cleaned.df, main="after adjustment 1")

#adjustment 2 - add NA
heartHealth.cleaned.df = heartHealth.cleaned.df[-c(106), ]
pairs20x(heartHealth.cleaned.df, main="after adjustment 2")

#adjustment 3
sdp = heartHealth.cleaned.df[44, "sbp"]
dbp = heartHealth.cleaned.df[44, "dbp"]
heartHealth.cleaned.df[44, "dbp"] = sdp
heartHealth.df[44, "sbp"] = dbp
pairs20x(heartHealth.cleaned.df, main="after adjustment 3")
```
```{r Eggs Exploration}

#remove any NA's
eggs.df = heartHealth.df[,c("ethnicity", "eggs")]
eggs.df = eggs.df[complete.cases(eggs.df[, "eggs"]),]

#box plot
plot(eggs.df$ethnicity, eggs.df$eggs, main="Ethnicity versus typical number of eggs consumed per week",
     xlab="Ethnicity", ylab="Number of eggs consumeed per week")
#pairs plot
pairs20x(eggs.df[c("ethnicity", "eggs")])

#fit model
eggs.fit = glm(eggs ~ ethnicity, data=eggs.df, family="poisson")
#anova 
anova(eggs.fit, test = "Chisq")
#summary 
summary(eggs.fit)

#confidence intervals - european baseline
100*(exp(confint(eggs.fit)) -1)
```

```{r Drinking}
#subset and remove any NA's
drinking.df = heartHealth.df[, c("drinkmaxday", "age", "sex")]

pairs20x(drinking.df, main="Pairs plot of age, sex and maximum number of \n alcoholic drinks consumed per day")

#fit model
drinks.fit = glm(drinkmaxday ~ age*sex, data=drinking.df, family="poisson")
summary(drinks.fit)

#fit model - remove interaction
drinks.fit = glm(drinkmaxday ~ age+sex, data=drinking.df, family="poisson")
#model summary
summary(drinks.fit)
#confidence intervals 
100*(exp(confint(drinks.fit))-1)

#plot part d
ages <- data.frame(age=seq(0,100), sex="M")
malePredictedValues <- predict(drinks.fit, newdata = ages)
malePredictedValues <- exp(malePredictedValues)
ages <- data.frame(age=seq(0,100), sex="F")
femalePredictedValues <- predict(drinks.fit, newdata = ages)
femalePredictedValues <- exp(femalePredictedValues)

plot(drinkmaxday ~ age, data = heartHealth.df,col = ifelse(heartHealth.df$sex == "M" , "blue", "red"), xlim=c(0, 100), 
     main="Age versus Maximum number of alcoholic drinks \n consumed per day subset by gender", xlab="Age", ylab="Maximum number of alcoholic drinks \n consumed per day")
lines(ages$age, femalePredictedValues, col="red")
lines(ages$age, malePredictedValues, col="blue")
legend("topright", legend=c("Female", "Male"),
       col=c("red", "blue"), lty=1)

#e
test1.df = data.frame(age=c(30,50), sex=c("M", "M"))
test1.predict = predict(drinks.fit, newdata = test1.df, type = "response")
test2.df = data.frame(age=c(40,40), sex=c("M", "F"))
test2.predict = predict(drinks.fit, newdata = test2.df, type = "response")
print(test1.predict)
print(test2.predict)
```

## Exercise

b) The response variable is continous and not a count/proportion type and thus a linear regression model is appropriate. 

c) We have evidence to show that age and smoking history are related to the number of hours typical number of hours of vigorous exercise undertaken in a week. However, we have no evidence that the relationship between age and the number of hours typical number of hours of vigorous exercise undertaken in a week, depended on smoking history.

Interpretation:
We estimate that 

```{r Exercise}

exercise.df = heartHealth.df[c("exermin", "exerhour", "exerday", "smoke", "age")]

#create exerHours field 
exerMinsInHours = exercise.df$exermin/60
exerHoursPerDay =  exercise.df$exerhour + exerMinsInHours
exerHoursPerWeek = exercise.df$exerday * exerHoursPerDay
exercise.df$exerHoursPerWeek = exerHoursPerWeek

#fit model
exer.fit = glm(exerHoursPerWeek ~ age*smoke, data=exercise.df, family="gaussian")
summary(exer.fit)
#fit model - remove interaction
exer.fit = glm(exerHoursPerWeek ~ age+smoke, data=exercise.df, family="gaussian")
summary(exer.fit)
confint(exer.fit)


#plot d 
pred.df <- data.frame(age=seq(0,100), smoke="Yes")
smokersPredictions <- predict(exer.fit, newdata = pred.df)
pred.df <- data.frame(age=seq(0,100), smoke="No")
nonsmokersPredictions <- predict(exer.fit, newdata = pred.df)

plot(exerHoursPerWeek ~ age, data = exercise.df, col = ifelse(exercise.df$smoke == "Yes" , "blue", "red"),xlim=c(0, 100), 
     main="Age versus number of hours exercised per week \n subset by smoking history", xlab="Age", 
     ylab = "Number of hours exercised per week")
lines(pred.df$age, nonsmokersPredictions, col="red")
lines(pred.df$age, smokersPredictions, col="blue")
legend("topright", legend=c("Non-Smokers", "Smokers"),
       col=c("red", "blue"), lty=1)

```

## Heart Attacks

```{r Heart Attacks}

heartattack.df = heartHealth.df[c("heartattack", "chol", "age")]

#fit model
heartattack.fit = glm(heartattack ~ age*chol, data=heartattack.df, family="binomial")
summary(heartattack.fit)

#fit model - remove interaction
heartattack.fit = glm(heartattack ~ age+chol, data=heartattack.df, family="binomial")
summary(heartattack.fit)

100*(exp(confint(heartattack.fit))-1)

#plot d
pred.df <- data.frame(age=seq(0,100), chol="High")
highCholPredictions <- predict(heartattack.fit, newdata = pred.df)
highCholPredictionsOdds = exp(highCholPredictions)
highCholPredictionsProb = highCholPredictionsOdds / (1 + highCholPredictionsOdds)

pred.df <- data.frame(age=seq(0,100), chol="Low")
lowCholPredictions <- predict(heartattack.fit, newdata = pred.df)
lowCholPredictionsOdds = exp(lowCholPredictions)
lowCholPredictionsProb = lowCholPredictionsOdds / (1 + lowCholPredictionsOdds)

plot(heartattack ~ age, data = heartattack.df, col = ifelse(heartattack.df$chol == "High" , "blue", "red"),xlim=c(0,100), ylim=c(0,1), 
     main="Age vs Probability of having a heart-attack subset by \n Cholestrol level", xlab="Age", ylab="Probability of having a heart attack")
lines(pred.df$age, lowCholPredictionsProb, col="red")
lines(pred.df$age, highCholPredictionsProb, col="blue")
legend("topleft", legend=c("Low Cholestrol", "High Cholestrol"),
       col=c("red", "blue"), lty=1)
```
