---
title: "Stats  330 Assignment 1"
author: "Hasnain Cheena"
date: "16/03/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
library(ggplot2)
```

## Data Cleaning

```{r Data Cleaning}
heartHealth.df = read.csv("hearthealth.csv")
heartHealthCleaned.df = read.csv("hearthealth_cleaned.csv")
heartHealthCleanedCondensed.df = heartHealthCleaned.df[c("height", "weight", "sbp", "dbp")]
heartHealthCondensed.df = heartHealth.df[c("height", "weight", "sbp", "dbp")]
pairs20x(heartHealthCondensed.df)
pairs20x(heartHealthCleanedCondensed.df)
```
From the pairs plot you can see that height has a single outlier (height = 15.1 metres).This height is not sensible and I propose the decimal point was inputted in the wrong place. This is because if we examine similar instances (females between 40-45 years old with similar other features) we can see their heights are between 1.53m and 1.65m. Therefore, I believe the outlier is a mistake and the height should be modified to 1.51m. 

From the pairs plot you can see that weight has a single outlier (weight = 13.4kg). The subject is a European male of height 1.75m, therefore a weight of 13.4kg is infeasible. I believe this value is an error and as such have decided to remove the instance. 

There's another outlier in the dbp. I believe the dbp and sbp values were swapped for this person upon data entry. Therefore, I have swapped the sbp and dbp values for this person. 

## Eggs Exploration
Firstly did some initial data exploration creating a box plot showing ethnicity versus number of eggs consumed per week. 
The plot showed a massive outlier for a Maori person that ate 60 eggs a week. 

Response variable is a discrete count therefore a Poisson regression model was used. 

#### Interpret Model Results
We have strong evidence to suggest that the expected number of eggs consumed per week by a Maori person is higher than a European person. 
We have strong evidence to suggest that the expected number of eggs consumed per week by a Polynesian person is higher than a European person.
We have no evidence to suggest that there is a difference in the expected number of eggs consumed per week between a European person and Other ethnicity person. 

We estimate that the expected number of eggs conusmed per week by a European person is 1.9 to 2.5 eggs.
We estimate that the expected number of eggs consumed per week by a Maori person is 3.3 to 5 times higher than an European person.
We estimate that the expected number of eggs consumed per week by a Polynesian person is 1.1 to 2.5 times higher than an European person. 
```{r Eggs Exploration}

#remove Na rows in eggs
  

#box plot
plot(heartHealthCleaned.df$ethnicity, heartHealthCleaned.df$eggs)

#fit model
eggs.fit = glm(eggs ~ ethnicity, data=heartHealthCleaned.df, family="poisson")

#model summary
summary(eggs.fit)
#confidence intervals 
exp(confint(eggs.fit))
```

## Drinking 
a) The explanatory variables that will be used in this case are: age, sex and drinkmaxday. 

b) A Poisson regression model would be most appropriate to fit. This is because the response variable drinkmaxday is a discrete count. 

c) Initially fit a model with interaction but the interaction was deemed insignificant so removed it. After removal both main effects became significant. 

#### Model Interpretation
We estimate that for each additional year of age (for both males and females) the expected number of the maximum number of alcoholic drinks consumed per day (within the last 3 months) decreases by 0.97 to 0.99 times. 
We estimate that ,for the same age, the expected number of of the maximum number of alcoholic drinks consumed per day (within the last 3 months) for male is 1.3 to 1.9 times higher than a female. 

####  Is there evidence to suggest that age and/or gender are somehow related to the maximum number of alcoholic drinks consumed per day (within the last 3 months)?
We have strong evidence to suggest that both age and gender are related to the maximum number of drinks consumed per day (within the last 3 months). However, there is no evidence that the relationship between age and the maximum number of drinks consumed per day (within the last 3 months) depended on gender (and vice versa). 


d) Plotted
  ii) The plot show us the expected value of maximum number of alcoholic drinks consumed per day (within the last 3 months) is higher for males than females for the same age. Further, the plot shows that as a person gets older the maximum number of alcoholic drinks consumed per day (within the last 3 months) is decreases in an expoential fashion for both males and females. Further this decreasing relationship is the same for both males and females. 
  
e) 
We estimate that the expected value of maximum number of alcoholic drinks consumed per day (within the last 3 months) for a 30 year old male is 8.5 drinks. In comparision we estimate that the expected value of maximum number of alcoholic drinks consumed per day (within the last 3 months) for a 50 year old male is 5.7 drinks. 

We estimate the that the expected value of maximum number of alcoholic drinks consumed per day (within the last 3 months) for a 40 year old male is 6.9 drinks. In comparision we estimate that the expected value of maximum number of alcoholic drinks consumed per day (within the last 3 months) for a 40 year old female is 4.4 drinks.
```{r Drinking}
#fit model
drinks.fit = glm(drinkmaxday ~ age*sex, data=heartHealthCleaned.df, family="poisson")
#model summary
summary(drinks.fit)

#fit model
drinks.fit = glm(drinkmaxday ~ age+sex, data=heartHealthCleaned.df, family="poisson")
#model summary
summary(drinks.fit)
#confidence intervals 
exp(confint(drinks.fit))

#plot part d
ages <- data.frame(age=seq(0,200), sex="M")
malePredictedValues <- predict(drinks.fit, newdata = ages)
malePredictedValues <- exp(malePredictedValues)
ages <- data.frame(age=seq(0,200), sex="F")
femalePredictedValues <- predict(drinks.fit, newdata = ages)
femalePredictedValues <- exp(femalePredictedValues)

plot(drinkmaxday ~ age, data = heartHealthCleaned.df,col = ifelse(heartHealthCleaned.df$sex == "M" , "black", "red"))
lines(ages$age, femalePredictedValues, col="red")
lines(ages$age, malePredictedValues, col="black")

plot(ages$age, femalePredictedValues, col="red", type="l")
lines(ages$age, malePredictedValues, col="black", type="l")


#e
test1.df = data.frame(age=c(30,50), sex=c("M", "M"))
test1.predict = exp(predict(drinks.fit, newdata = test1.df))
test2.df = data.frame(age=c(40,40), sex=c("M", "F"))
test2.predict = exp(predict(drinks.fit, newdata = test2.df))
print(test1.predict)
print(test2.predict)
```

## Exercise

a) In this situation the explanatory variables are age and smoke (smoking history). The response variable is exerHoursPerWeek. This variable is a combination of exermin, exerhour and exerday. The method of calculation is shown below. 

b) The response variable is continous and not a count/proportion type and thus a linear regression model is appropriate. 

c) We have evidence to show that age and smoking history are related to the number of hours typical number of hours of vigorous exercise undertaken in a week. However, we have no evidence that the relationship between age and the number of hours typical number of hours of vigorous exercise undertaken in a week, depended on smoking history.

Interpretation:
We estimate that 

```{r Exercise}
exerMinsInHours = heartHealthCleaned.df$exermin/60
exerHoursPerDay =  heartHealthCleaned.df$exerhour + exerMinsInHours
exerHoursPerWeek = heartHealthCleaned.df$exerday * exerHoursPerDay
heartHealthCleaned.df$exerHoursPerWeek = exerHoursPerWeek

#fit model
exer.fit = glm(exerHoursPerWeek ~ age+smoke, data=heartHealthCleaned.df, family="gaussian")
#model summary
summary(exer.fit)
#confidence intervals
confint(exer.fit)
```

## Heart Attacks

```{r Heart Attacks}


```
