---
title: "326 Assignment 2"
author: "Hasnain Cheena"
date: "30/03/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
```

```{r Data Input, echo=FALSE}
CO2.df = read.table('Cape Grim.CO2.quarterly.txt', header=TRUE) 

#create time series object
CO2.fit.ts = ts(CO2.df[1:75, ], frequency = 4, start=c(2000,1))
CO2.pred.ts = ts(CO2.df[76:79, ], frequency = 4, start=c(2018,4))
CO2.full.ts = ts(CO2.df$CO2, frequency=4, start=c(2000,1))

#setup Time
Time = 1:79
reduced.Time = 1:75
#observations 1 to 75 used to fit the model
#observations 76 to 79 used to test models
```


## Question 1
```{r Q1 - HW Fit}
HW.fit = HoltWinters(CO2.fit.ts)
HW.fit

HW.pred = predict(HW.fit,n.ahead=4)
HW.pred

HW.RMSEP = sqrt(1/4*sum((CO2.pred.ts-HW.pred)^2))
HW.RMSEP

plot(HW.fit, HW.pred, main="Cape Grim CO2 - Holt Winters", xlab="Year", ylab="Observed/Fitted CO2 (ppm)")
```
The plot of the Holt-Winters model shows the model is good fit. This is becaue the model (red line) is very close to the actual observations (black line) with very litte white space between the lines.  

The RMSEP indicates that, on average, the prediction error is 0.22 ppm.

## Question  2 

```{r Q2 - Plot}
#calculate and extract seasonal estimates
decomp.stl.CO2 = stl(CO2.fit.ts, s.window="periodic")
#de-seasonalize 
stl.CO2.ts = CO2.fit.ts - decomp.stl.CO2$time.series[,1]
quick.fit = lm(CO2.df$CO2~Time)

plot(stl.CO2.ts, main="STL seasonally adjusted CO2 Cape Grim", xlab="Year", ylab="CO2 (ppm)")
abline(quick.fit, lty=2)
abline(v=50, lty=3)

```
The plot of the seasonal-trend-lowess seasonally adjusted series shows a reasonably increasing linear trend. 


```{r Q2 - Seasonal Estimates}
#seasonal estimates
decomp.stl.CO2$time.series[1:4,1]
```
The seasonal estimates for Quarter 1 & Quarter 2 are similar negative values (below the trend). In comparision Quarter 3 & Quarter 4 are positive and above the trend. 
Quarter 3 is the highest (0.48) while Quarter 1 is the lowest (-0.39). 

```{r Q2 - Model Fit 1, echo=FALSE}

Time.break = c(rep(0,49), Time[50:75]-Time[50])

CO2.break.fit3 = lm(stl.CO2.ts[-1] ~ reduced.Time[-1]+Time.break[-1] + stl.CO2.ts[-75])
plot.ts(residuals(CO2.break.fit3), main="Residual Series") #residuals are 0 mean with constant variance 
acf(residuals(CO2.break.fit3), main="ACF Plot of Residual Series") #still significant lag at lag(1) - best we can do so far
normcheck(CO2.break.fit3, shapiro.wilk = TRUE) #check normality - seems normal but need more information about 

stl.seasonal.estimates = decomp.stl.CO2$time.series
```

```{r Q2 - Model Fit}
#model summary
summary(CO2.break.fit3)

#forecast 2018 Q4
t76.sa.pred = CO2.break.fit3$coefficients[1] + CO2.break.fit3$coefficients[2] * 76 + CO2.break.fit3$coefficients[3] * 26 + CO2.break.fit3$coefficients[4]  * stl.CO2.ts[75]
t76.pred = t76.sa.pred + stl.seasonal.estimates[4,1]

#forecast 2019 Q1
t77.sa.pred = CO2.break.fit3$coefficients[1] + CO2.break.fit3$coefficients[2] * 77 + CO2.break.fit3$coefficients[3] * 27 + CO2.break.fit3$coefficients[4]  * t76.sa.pred
t77.pred = t77.sa.pred + stl.seasonal.estimates[1,1]

#forecast 2019 Q2
t78.sa.pred = CO2.break.fit3$coefficients[1] + CO2.break.fit3$coefficients[2] * 78 + CO2.break.fit3$coefficients[3] * 28 + CO2.break.fit3$coefficients[4]  * t77.sa.pred
t78.pred = t78.sa.pred + stl.seasonal.estimates[2,1]

#forecast 2019 Q3
t79.sa.pred = CO2.break.fit3$coefficients[1] + CO2.break.fit3$coefficients[2] * 79 + CO2.break.fit3$coefficients[3] * 29 + CO2.break.fit3$coefficients[4]  * t78.sa.pred
t79.pred = t79.sa.pred + stl.seasonal.estimates[3,1]

results.df = data.frame("Time"=c("2018 Q4", "2019 Q1", "2019 Q2", "2019 Q3"),
                        "Seasonally Adjusted"=c(t76.sa.pred,t77.sa.pred,t78.sa.pred,t79.sa.pred), 
                        "Predictions"=c(t76.pred,t77.pred,t78.pred,t79.pred))

results.df

#calculte RMSEP
stl.pred.ts = ts(results.df$Predictions, start=c(2018,4), frequency = 4)

#RMSEP
STL.RMSEP = sqrt(1/4*sum((CO2.pred.ts-stl.pred.ts)^2))
STL.RMSEP
```

## Question 3 - Technical Notes
The seasonal estimates show that the CO2 concentration is below the overall trend for Quarter 1 & Quarter 2 with Quarter 1 being the lowest (-0.39) and the CO2 concenetration is above the overall trend for Quarter 3 & Quarter 4, with Quarter 3 being the highest (0.48).

The plot of the seasonally adjusted series shows a change in the slope/break in the trend around 2012 Quarter 2 (t=50), with the trend after t=50 being slightly higher than previously. 


The final model fitted to the seasonal trend lowess seasonally adjusted series included a break-in trend term and was also corrected for auto-correlation. 

For the final model the residual series appears to be random scatter about 0. The plot of the autocorrelation function shows a significant autocorrelation at positive lag(1) and lag(11) and significant negative autocorrelation at lag(16). 
The residuals appear to be normally distributed (Shaprio-Wilk P-value = 0.83) with a 
Since we have autocorrelation in the Residual Series, the estimate of the error standard deviation and variance of the estimates and will be underestimated. Further, $R^2$ will be overestimated and the t-statistics and F-statistic will not be valid. The model $CO2.break.fit3$ does not satisfy the assumptions for a linear regression model. 

We have strong evidence against the hypothesis that the coefficient associated with the time variable is 0 (P-value \approx 0.0001). Further, we have strong against the hypothesis that the coefficient associated with the break-in trend time variable is 0 (P-value \approx 0.0005). Additionally, we have very strong evidence against the hypothesis of no autocorrelation (P-value \approx 0).

The F-statistic provides extremely strong evidence against the hypothesis that none of the variables are related to the seasonally adjusted CO2 concentration (P-value \approx 0). The multiple $R^2$ is 0.9997 indicating that 99% of the variation in the seasonally adjusted CO2 concentration is explained by our model. Prediction wil be unreliable as the assumptions are not satisfied. 

## Question 4
```{r Q4, echo=FALSE}
Time = 1:79
Time.break = c(rep(0,49), Time[50:79]-Time[50])
CO2.df = read.table('Cape Grim.CO2.quarterly.txt', header=TRUE) 
CO2.full.ts = ts(CO2.df$CO2, frequency=4, start=c(2000,1))
decomp.stl.CO2 = stl(CO2.full.ts, s.window="periodic")
#de-seasonalize 
stl.CO2.ts = CO2.full.ts - decomp.stl.CO2$time.series[,1]
stl.seasonal.estimates = decomp.stl.CO2$time.series
```


```{r Q4 - Model Fit}
#best predicting model - STL model

#rerun the model using all avaliable data
CO2.break.fit4 = lm(stl.CO2.ts[-1] ~ Time[-1]+Time.break[-1] + stl.CO2.ts[-79])

#model summary
summary(CO2.break.fit4)
```

```{r Q4 - Forecasts}
#forecast 2019 Q4
t80.sa.pred = CO2.break.fit4$coefficients[1] + CO2.break.fit4$coefficients[2] * 80 + CO2.break.fit4$coefficients[3] * 30 + CO2.break.fit4$coefficients[4]  * stl.CO2.ts[79]
t80.pred = t80.sa.pred + stl.seasonal.estimates[4,1]

#forecast 2020 Q1
t81.sa.pred = CO2.break.fit4$coefficients[1] + CO2.break.fit4$coefficients[2] * 81 + CO2.break.fit4$coefficients[3] * 31 + CO2.break.fit4$coefficients[4]  * t80.sa.pred
t81.pred = t81.sa.pred + stl.seasonal.estimates[1,1]

#forecast 2020 Q2
t82.sa.pred = CO2.break.fit4$coefficients[1] + CO2.break.fit4$coefficients[2] * 82 + CO2.break.fit4$coefficients[3] * 32 + CO2.break.fit4$coefficients[4]  * t81.sa.pred
t82.pred = t82.sa.pred + stl.seasonal.estimates[2,1]

#forecast 2020 Q3
t83.sa.pred = CO2.break.fit4$coefficients[1] + CO2.break.fit4$coefficients[2] * 83 + CO2.break.fit4$coefficients[3] * 33 + CO2.break.fit4$coefficients[4]  * t82.sa.pred
t83.pred = t83.sa.pred + stl.seasonal.estimates[3,1]

#convert to ts object
stl.pred.df = data.frame(CO2=c(t80.pred,t81.pred,t82.pred,t83.pred))
stl.pred.ts = ts(stl.pred.df, start=c(2019,4), frequency = 4)
stl.pred.ts
```

The seasonal-trend-lowess model estimates for the full timeframe model ($CO2.break.fit4$) are similar to those from the reduced timeframe model ($CO2.break.fit3$). 
