---
title: 'Lab 1: Wikipedia searches'
author: "Hasnain Cheena 19041106 hche737"
date: "5 August 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
```

Task description and data for candidates applying to be a Data Analyst in the [Discovery department](https://www.mediawiki.org/wiki/Wikimedia_Discovery) at [Wikimedia Foundation](https://wikimediafoundation.org/wiki/Home).

## Background

Discovery (and other teams within the Foundation) rely on *event logging* (EL) to track a variety of performance and usage metrics to help us make decisions. Specifically, Discovery is interested in:

- *clickthrough rate*: the proportion of search sessions where the user clicked on one of the results displayed
- *zero results rate*: the proportion of searches that yielded 0 results

and other metrics outside the scope of this task. EL uses JavaScript to asynchronously send messages (events) to our servers when the user has performed specific actions. In this task, you will analyze a subset of our event logs.


## Data

The dataset comes from a [tracking schema](https://meta.wikimedia.org/wiki/Schema:TestSearchSatisfaction2) that we use for assessing user satisfaction. Desktop users are randomly sampled to be anonymously tracked by this schema which uses a "I'm alive" pinging system that we can use to estimate how long our users stay on the pages they visit. The dataset contains just a little more than a week of EL data.

| Column          | Value   | Description                                                                       |
|:----------------|:--------|:----------------------------------------------------------------------------------|
| uuid            | string  | Universally unique identifier (UUID) for backend event handling.                  |
| timestamp       | integer | The date and time (UTC) of the event, formatted as YYYYMMDDhhmmss.                |
| session_id      | string  | A unique ID identifying individual sessions.                                      |
| group           | string  | A label ("a" or "b").                                     |
| action          | string  | Identifies in which the event was created. See below.                             |
| checkin         | integer | How many seconds the page has been open for.                                      |
| page_id         | string  | A unique identifier for correlating page visits and check-ins.                    |
| n_results       | integer | Number of hits returned to the user. Only shown for searchResultPage events.      |
| result_position | integer | The position of the visited page's link on the search engine results page (SERP). |

The following are possible values for an event's action field:

- **searchResultPage**: when a new search is performed and the user is shown a SERP.
- **visitPage**: when the user clicks a link in the results.
- **checkin**: when the user has remained on the page for a pre-specified amount of time.

### Example Session

|uuid                             |      timestamp|session_id       |group |action           | checkin|page_id          | n_results| result_position|
|:--------------------------------|:--------------|:----------------|:-----|:----------------|-------:|:----------------|---------:|---------------:|
|4f699f344515554a9371fe4ecb5b9ebc | 20160305195246|001e61b5477f5efc |b     |searchResultPage |      NA|1b341d0ab80eb77e |         7|              NA|
|759d1dc9966353c2a36846a61125f286 | 20160305195302|001e61b5477f5efc |b     |visitPage        |      NA|5a6a1f75124cbf03 |        NA|               1|
|77efd5a00a5053c4a713fbe5a48dbac4 | 20160305195312|001e61b5477f5efc |b     |checkin          |      10|5a6a1f75124cbf03 |        NA|               1|
|42420284ad895ec4bcb1f000b949dd5e | 20160305195322|001e61b5477f5efc |b     |checkin          |      20|5a6a1f75124cbf03 |        NA|               1|
|8ffd82c27a355a56882b5860993bd308 | 20160305195332|001e61b5477f5efc |b     |checkin          |      30|5a6a1f75124cbf03 |        NA|               1|
|2988d11968b25b29add3a851bec2fe02 | 20160305195342|001e61b5477f5efc |b     |checkin          |      40|5a6a1f75124cbf03 |        NA|               1|

This user's search query returned 7 results, they clicked on the first result, and stayed on the page between 40 and 50 seconds. (The next check-in would have happened at 50s.)



# Task

```{r Load Data}
df = read_csv("events_log.csv")

#convert timestamp to datetime format
df = df %>%
  mutate(datetime=parse_date_time(timestamp, c("ymd HMS")))
```
## Question 1

### 1a) What is the overall clickthrough rate? 

This section was computed by finding the number sessions with one or more one clicks. Then dividing the number of sessions with one or more clicks by the total number of sessions that occurred. The result is the overall clickthrough rate. 

```{r Task 1a}

overall_click_through_stats = df %>%
  count(action, session_id, name="action_count") %>% 
  filter(action == "visitPage") %>%
  mutate(visit_page_flag =  action_count > 0) %>%
  summarise(total_click_throughs = sum(visit_page_flag), 
            total_sessions = length(unique(df$session_id)), 
            overall_click_through_rate = total_click_throughs/total_sessions * 100) 


round(overall_click_through_stats$overall_click_through_rate,1)       
                          
```
The overall clickthrough rate is 38.9%. 

### 1b) How does the overall clickthrough rate vary between the groups?

In this section, the clickthrough rates were examined on a group level. The number of sessions observed per group was calculated. Then the number of sessions with one or more clicks was calculated on a group level and divided by the number of sessions per group. 

```{r Task 1b}
#find number of sessions per group
number_of_sessions_per_group = df %>%
  distinct(session_id, group) %>%
  count(group, name="number_of_sessions")

#find number of clickthroughs per group
number_sessions_with_clickthrough_per_group = df %>%
  count(action, session_id, group, name="action_count") %>%
  filter(action == "visitPage") %>%
  mutate(visit_page_flag =  action_count > 0) %>%
  group_by(group) %>%
  summarise(total_click_throughs = sum(visit_page_flag))


# overall_click_through_rate_per_group
overall_click_through_rate_per_group = inner_join(number_of_sessions_per_group, number_sessions_with_clickthrough_per_group, by="group") %>%
  summarise(group=group, overall_click_through_rate = total_click_throughs/number_of_sessions * 100) 

overall_click_through_rate_per_group
                          
```
The clickthrough rate greatly varies per group as can be seen from the results above. 
Group 'a' has a clickthrough rate of 67.0% versus group 'b' only has a clickthrough rate of 17.5%. 

## Question 2
### Which results do people tend to try first? 
Note that this section was computed by first restricting the problem to sessions with a single search (as allowed in the lab sheet). Then the positions of the first clicked result were counted. 

```{r Task 2}
#which results do people tend to try first
#restrict to sessions with a single search

session_ids_single_search = df %>% 
  select(session_id, action) %>%
  group_by(session_id) %>%
  count(action, name="count_of_action") %>%
  filter(action == "searchResultPage" &  count_of_action==1) %>%
  select(session_id)

#join to filter and only get single search sesssions
df_restricted = inner_join(df, session_ids_single_search, by="session_id")

#find the distribution of result positions of clicked results
distribution_of_clicked_result_positions = df_restricted %>% 
  arrange(datetime) %>%
  filter(action == "visitPage") %>%
  select(datetime, session_id, action, result_position) %>%
  group_by(session_id) %>%
  summarise(position_of_first_result_clicked = first(result_position)) %>%
  ungroup() %>%
  count(position_of_first_result_clicked, name = "number_of_clicks")
  

distribution_of_clicked_result_positions
```
It can be seen that majority people tend to try the 1st result on the page. 