---
title: "326 A1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Read in data
```{r data}
aud.df = read.csv("Data\\NZD To AUD.csv")
sunspot.df = read.csv("Data\\Monthly Sunspot.csv")
xrp.df = read.csv("Data\\XRP Daily Price.csv")
temp.df = read.csv("Data\\GlobalLandTemperaturesByMajorCity.csv")
milkprod.df = read.csv("Data\\Milk Production.csv", nrows=58)
milkprodLong.df = read.csv("Data\\2002 NZ Milk production data.csv")
```

## Question 1
```{r NZD AUD rate}
#plot time vs exchange rate
aud.ts = ts(aud.df$Rate, frequency=12, start=c(1990,1))
plot(aud.ts, main="AUD to NZD Exchange Rate (Jan 1990 - March 2020)", 
xlab="Year",ylab="Monthly average exchange rate")
```
## Question 2 - Find a Stationary Time Series.
Found from: https://coinmarketcap.com/currencies/xrp/
```{r Change in XRP value}
#plot time vs change in xrp price 
#use all values apart from first row as wrong 
changeInXrp.ts = ts(xrp.df$Change.in.average.daily.price[2:367], frequency=365, start=c(2019,2))

plot(changeInXrp.ts, main="Change in daily average price of XRP from Jan 2019 to Jan 2020",
     xlab="Year", ylab="Change in XRP value (USD)")


```


## Question 3 - Find a Time Series that has a seasonal component but no trend or cycle.
Found from: http://www-eio.upc.edu/~pau/cms/rdata/datasets.html - called sunspot.month
```{r Sunspot Number}
#Mike: Seasonal component - every 10 year there is a peak
#plot time vs exchange rate
sunspot.ts = ts(sunspot.df$sunspot.month, frequency=12, start=c(1749,1))
plot(sunspot.ts, main="Monthly Sunspot Numbers (1749 to 2013)", 
xlab="Year", ylab="Sunspot Numbers")
```

## Question 3 - Find a Time Series that has a seasonal component but no trend or cycle.
Found from: https://www.kaggle.com/marvintherain/average-monthly-temp/data
```{r Mean Monthly Temperature}
#extract wuhan 
temp.df = temp.df[temp.df$City == 'Wuhan', c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#extract 2000 onwards
temp.df = temp.df[1909:2072, c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#add time axis
temp.df["Time"] = temp.df$dt

#plot time vs temp
temp.ts = ts(temp.df$AverageTemperature, frequency=12, start=c(2000,1))
plot(temp.ts)
```
## Question  - Find a Time Series that has a seasonal component but no trend or cycle. 
Found from: https://www.dcanz.com/UserFiles/DCANZ/File/191119%20New%20Zealand%20Milk%20Production%2019Nov2019.pdf
```{r Milk Prod}
milkProd.ts = ts(milkprod.df$Milk.Production, freq=12, start=c(2015,1))
plot(milkProd.ts)
```

## Question  - Find a Time Series that has a reasonably linear trend and has a seasonal component. 
Found from: https://www.dcanz.com/about-the-nz-dairy-industry/
```{r Milk Prod Long term}
milkProdLong.ts = ts(milkprodLong.df$X.000.kg.Milksolids, freq=12, start=c(2003,6))
plot(milkProdLong.ts)
```