---
title: "Stats 326 Assignment 1"
author: 'Name: Hasnain Cheena, ID:190411106'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Read in data
```{r Data Sources}
aud.df = read.csv("Data/NZD To AUD.csv")
xrp.df = read.csv("Data/XRP Daily Price.csv")
temp.df = read.csv("Data/GlobalLandTemperaturesByMajorCity.csv")
alc.df = read.csv("Data/NZAlcoholConsumption.csv")
gold.df = read.csv("Data/Gold Prices.csv")
```

## Question 1 - Find a Time Series that exhibits cycles
```{r NZD AUD rate}
#plot time vs exchange rate
aud.ts = ts(aud.df$Rate, frequency=12, start=c(1990,1))
plot(aud.ts, main="Monthly Average AUD to NZD Exchange Rate (Jan 1990 - March 2020)", 
xlab="Year",ylab="Monthly average exchange rate")

#stl decomposition
plot(stl(aud.ts, s.window="periodic"))
```
The data was found from: https://www.ofx.com/en-us/forex-news/historical-exchange-rates/monthly-average-rates/


The series shows strong cyclical behaviour with reasonably large random variation. There is no clear trend or seasonal behaviour. 
The range of the monthly average exchange rate is between 0.70 AUD to 0.98 AUD. 


## Question 2 - Find a Stationary Time Series.
Found from: https://coinmarketcap.com/currencies/xrp/
```{r Change in XRP value}
#plot time vs change in xrp price 
#use all values apart from first row as wrong 
changeInXrp12.ts = ts(xrp.df$Change.in.average.daily.price[2:367], frequency=12, start=c(2019,2))
changeInXrp.ts = ts(xrp.df$Change.in.average.daily.price[2:367], frequency=365, start=c(2019,2))

plot(changeInXrp.ts, main="Change in daily average price of XRP from Jan 2019 to Jan 2020",
     xlab="Year", ylab="Change in XRP value (USD)")

plot(stl(changeInXrp12.ts, s.window="periodic"))
```

Data was found from: https://coinmarketcap.com/currencies/xrp/
The series above shows no seasonality, trend or cycles. It is mean-centered at 0 and has reasonably constant variance. Thus it is a stationary time series. The large spikes in the plot indicate that there were large price fluctuations between June 2019 and August 2019.

## Question 2 - Find a Stationary Time Series.
Found from: https://www.gold.org/goldhub/data/gold-prices
```{r Gold prices}
gold.ts = ts(gold.df$Deviation[242:373], frequency=12, start=c(1990, 1))
plot(gold.ts, main="Deviations in the average monthly price of Gold (Jan 1990 to Dec 2000", ylab="USD", xlab="Time")

plot(stl(gold.ts, s.window = "period"))
```

Data was found from: https://www.gold.org/goldhub/data/gold-prices
The time series above shows no seasonality, trend or cycles. The plot displays that the time series is mean-centered at 0 and has reasonably constant variance. Thus it is a stationary time series and can be further classified as white noise. 

## Question 3 - Find a Time Series that has a seasonal component but no trend or cycle.
Found from: https://www.kaggle.com/marvintherain/average-monthly-temp/data
```{r Mean Monthly Temperature}
#extract wuhan 
temp.df = temp.df[temp.df$City == 'Wuhan', c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#extract 2000 onwards
temp.df = temp.df[1909:2072, c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#add time axis
temp.df["Time"] = temp.df$dt

#plot time vs temp
temp.ts = ts(temp.df$AverageTemperature, frequency=12, start=c(2000,1))
plot(temp.ts, main="Monthly Temperature in Wuhan, China (2001 to 2013)", 
xlab="Year", ylab="Temperature (Celcius)")

#stl decomposition
plot(stl(temp.ts, s.window="periodic"))

```
Data was found from: https://www.kaggle.com/marvintherain/average-monthly-temp/data

The series appears to have a strong seasonal component, although their is a large seasonal trough in 2008 and 2011. 
The annual seasonal pattern, with January having the lowest temperature and July having the highest. There does not appear to be any trend or cyclical behaviour. 
The range of the Temperature (in Celcius) is between 1 degree and 30 degrees. 

## Question 4 - Find a Time Series that has a reasonably linear trend and has a seasonal component. 
Found from: https://new.censusatschool.org.nz/resource/time-series-data-sets-2013/
```{r NZ Alcohol Consumption}
#extract years 2002 to 2006
alc.ts = ts(alc.df$TotalWine[9:29], freq=4, start=c(2002,1))
plot(alc.ts, main="Quarterly Total Wine Consumption in New Zealand (2002 to 2007)", 
xlab="Year", ylab="Wine consumption (Litres)")

#stl decomposition
plot(stl(alc.ts, s.window="periodic"))
```
Data found from: Found from: https://new.censusatschool.org.nz/resource/time-series-data-sets-2013/

The series above shows that the time series has a strong seasonal component and an reasonably linear increasing trend. The seasonal pattern implies that quarter 1 has the lowest quarterly value while quarter 4 has the highest quarterly value. Furthermore, there is a much lower seasonal trough in 2003. 
The range of the data appears to be between 1 and 4. 

## Question 5  
```{r Cape Grim CO2 Concentration}

#load data
capeGrim.df = read.table("Data\\CapeGrim CO2 2000.1 - 2019.9.txt",header = TRUE, sep = "\t")

#plot time series from January 2000 to September 2019
capeGrim.ts = ts(capeGrim.df$CO2, freq=12, start=c(2000,1))
plot(capeGrim.ts, main="Average Monthly CO2 Concentration at Cape Grim (Jan 2000 to Sept 2019)", xlab="Year", ylab="CO2 Concentration")

#convert data into average concentration for each quarter from 2000 to 2019.3
capeGrimQuarterly.ts = aggregate(capeGrim.ts,  nfrequency = 4, FUN=mean)
#plot the quarterly series
plot(capeGrimQuarterly.ts, main="Average Quarterly CO2 Concentration at Cape Grim (Jan 2000 to Sept 2019)", xlab="Year", ylab="CO2 Concentration")

#stl
plot(stl(capeGrimQuarterly.ts, s.window="periodic"))
```

The series shows a reasonably linear increasing trend in CO2 concentration at Cape Grim. Further, the series contains seasonal behaviour. 

The seasonal pattern shows a quarter 1 being the lowest quarterly value and quarter 3 with the highest quarerly value in CO2 concentration annually. 
The range of the CO2 concentration varies from 350 ppm and 450 ppm. 
