---
title: "Stats 326 Assignment 1"
author: 'Name: Hasnain Cheena, ID:190411106'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Sources for Q1 to Q4
```{r Data Sources}
aud.df = read.csv("Data/NZD To AUD.csv")
temp.df = read.csv("Data/GlobalLandTemperaturesByMajorCity.csv")
alc.df = read.csv("Data/NZAlcoholConsumption.csv")
gold.df = read.csv("Data/Gold Prices.csv")
```

## Question 1 - Find a Time Series that exhibits cycles
```{r NZD AUD rate}
#plot time vs exchange rate
aud.ts = ts(aud.df$Rate, frequency=12, start=c(1990,1))
plot(aud.ts, main="Monthly AUD/NZD Exchange Rate (Jan 1990 - March 2020)", 
xlab="Year",ylab="AUD")
```
The data was found from: https://www.ofx.com/en-us/forex-news/historical-exchange-rates/monthly-average-rates/

The series shows cyclical behaviour with reasonably large random variation. There are both major and minor cycles. There is no clear trend or seasonal behaviour. The range of the monthly average exchange rate is between $0.71 AUD to $0.98 AUD. 

## Question 2 - Find a Stationary Time Series.
Found from: https://www.gold.org/goldhub/data/gold-prices
```{r Gold prices}
gold.ts = ts(gold.df$Deviation[242:373], frequency=12, start=c(1990, 1))
plot(gold.ts, main="Deviations in monthly price of Gold (Jan 1990 to Dec 2000)", ylab="Change in Gold Price (USD)", xlab="Year")
```

Data was found from: https://www.gold.org/goldhub/data/gold-prices

The time series above is centred around 0 and has reasonably constant variance. Furthermore, the series has no trend, cycles or seasonality.
The range Thus it is a stationary time series. The range of the change in gold price is between -$39.2 USD and $44.2 USD. 

## Question 3 - Find a Time Series that has a seasonal component but no trend or cycle.
Found from: https://www.kaggle.com/marvintherain/average-monthly-temp/data
```{r Mean Monthly Temperature}
#extract wuhan 
temp.df = temp.df[temp.df$City == 'Wuhan', c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#extract 2000 onwards
temp.df = temp.df[1909:2072, c("dt", "AverageTemperature")]
rownames(temp.df) <- 1:nrow(temp.df) #reindex rows 
#add time axis
temp.df["Time"] = temp.df$dt

#plot time vs temp
temp.ts = ts(temp.df$AverageTemperature, frequency=12, start=c(2000,1))
plot(temp.ts, main="Monthly Temperature in Wuhan, China (Jan 2000 to Oct 2013)", 
xlab="Year", ylab="Temperature (Celcius)")
```
Data was found from: https://www.kaggle.com/marvintherain/average-monthly-temp/data

The series appears to have a strong seasonal component, although there are larger seasonal troughs in 2008 and 2011. 
The typical annual seasonal pattern, starts with January having the lowest temperature and July having the highest. There does not appear to be any trend or cyclical behaviour. The range of the Temperature (in Celcius) is between 1.5 degrees and 31.2 degrees. 

## Question 4 - Find a Time Series that has a reasonably linear trend and has a seasonal component. 
```{r NZ Alcohol Consumption}
#extract years 2002 to 2006
alc.ts = ts(alc.df$TotalWine[9:29], freq=4, start=c(2002,1))
plot(alc.ts, main="Quarterly Wine Consumption in New Zealand (Jan 2002 to Jan 2007)", 
xlab="Year", ylab="Wine consumption (Millions of litres)")
```
Data found from: Found from: https://new.censusatschool.org.nz/resource/time-series-data-sets-2013/

The series above shows that the time series has a strong seasonal component and an reasonably linear increasing trend. The seasonal pattern, once trend is considered, shows that typically quarter 1 has the lowest quarterly value while quarter 4 has the highest quarterly value. Furthermore, there is a much lower seasonal trough in 2003. The range of wine consumption (in millions of litres) is between 1.6 million litres and 3.3 million litres. 

## Question 5  
```{r Cape Grim CO2 Concentration}

#load data
capeGrim.df = read.table("Data/CapeGrim CO2 2000.1 - 2019.9.txt",header = TRUE, sep = "\t")

#plot time series from January 2000 to September 2019
capeGrim.ts = ts(capeGrim.df$CO2, freq=12, start=c(2000,1))
plot(capeGrim.ts, main="Monthly CO2 Concentration at Cape Grim (Jan 2000 to Sept 2019)", xlab="Year", ylab="CO2 Concentration (ppm)")

#convert data into average concentration for each quarter from 2000 to 2019.3
capeGrimQuarterly.ts = aggregate(capeGrim.ts,  nfrequency = 4, FUN=mean)
#plot the quarterly series
plot(capeGrimQuarterly.ts, main="Quarterly CO2 Concentration at Cape Grim (Jan 2000 to Sept 2019)", xlab="Year", ylab="CO2 Concentration (ppm)")
```

The series shows a reasonably linear increasing trend in CO2 concentration at Cape Grim. Further, the series displays seasonal behaviour. 
The seasonal pattern shows typically quarter 1 has the lowest quarterly value and quarter 4 has the highest quarterly value, once you account for the lineat trend.  The range of the CO2 concentration varies from 366.1 ppm and 408.6 ppm. 
