---
title: "Stats 326 - Assignment 5"
author: "Hasnain Cheena"
date: "27/05/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
```

```{r Data Setup, include=FALSE}
CO2.df = read.table('Cape Grim.CO2.quarterly.txt', header=TRUE) 

#create time series object
red.CO2.ts = ts(CO2.df[1:75, ], frequency = 4, start=c(2000,1))
pred.CO2.ts = ts(CO2.df[76:79, ], frequency = 4, start=c(2018,4))
full.CO2.ts = ts(CO2.df$CO2, frequency=4, start=c(2000,1))

#setup Time
Time = 1:79
reduced.Time = 1:75
#observations 1 to 75 used to fit the model
#observations 76 to 79 used to test models

#quarterly factor
Quarter = factor(rep(1:4,20)[-80])
reduced.Quarter = Quarter[1:75]

#time break setup
reduced.Time.break = c(rep(0,49), Time[50:75]-Time[50])
```

## Question 1

```{r Q1 - Fit}
#best fitting SARIMA
sarima.fit = arima(red.CO2.ts,order=c(0,1,1),
seasonal=list(order=c(0,1,1),period=4))
sarima.fit
```
```{r Q1 - Plot, fig.width=10, fig.height=6}
plot.ts(residuals(sarima.fit),main="Residual Series")
acf(residuals(sarima.fit))
normcheck(residuals(sarima.fit), shapiro.wilk = T)
```
The Residual Series show reasonably random scatter about 0, although there are two large negative residuals (2001 Quarter 1 and 2009 Quarter 2). The autocorrelation function plot of the Residual Series shows a significant lag at lag 11. This is an unsual lag to be significant in quarterly data. Furthermore, the residuals seems reasonably normally distributed. Therefore, all model assumptions are satifised. 
Normality is okay with CLT though?

```{r Q1 - Predictions}
#predictions
sarima.pred = predict(sarima.fit,n.ahead=4)
sarima.pred
#actual
pred.CO2.ts
#RMSEP
RMSEP.sarima = sqrt(1/4*sum((pred.CO2.ts-sarima.pred$pred)^2))
RMSEP.sarima
```

The $SARIMA(0,1,1)\times(0,1,1)_4$ model had an RMSEP of 0.33 ppm. The best predicting model from previous assginments was the seasonal-trend-lowess seasonally adjusted model as it had the lowest RMSEP (0.2 ppm). Therefore, the SARIMA model is not better predicting the the STL model has SARIMA has a higher RMSEP. 

## Question 2

```{r Q1 - Fit All}
#best fitting SARIMA
sarima.fit.full = arima(full.CO2.ts,order=c(0,1,1),
seasonal=list(order=c(0,1,1),period=4))
sarima.fit.full
```
Model in backshift notation:
$$ (1-B)(1-B^4)y_t = (1+\alpha_1B)(1+A_1B^4)\varepsilon_t$$ 
$$(1-B-B^4+B^5)y_t  = (1+\alpha_1B + A_1B^4 + \alpha_1A_1B^5)\varepsilon_t$$ 
$$y_t - y_{t-1} - y_{t-4} + y_{t-5} = \varepsilon_t + \alpha_1\varepsilon_{t-1} + A_1\varepsilon_{t-4} +  \alpha_1A_1\varepsilon_{t-5}$$
$$ y_t = y_{t-1} + y_{t-4} - y_{t-5} + \varepsilon_t + \alpha_1\varepsilon_{t-1} + A_1\varepsilon_{t-4} +  \alpha_1A_1\varepsilon_{t-5}$$ 
$$ y_t = y_{t-1} + y_{t-4} - y_{t-5} + \varepsilon_t + 0.5614\varepsilon_{t-1} - 0.8524\varepsilon_{t-4} - 0.4785\varepsilon_{t-5}$$
#### Predictions

Prediction 2019 Q4: $$ y_{t+1} = y_{t} + y_{t-3} - y_{t-4} + \varepsilon_{t+1} + 0.5614\varepsilon_{t} - 0.8524\varepsilon_{t-3} - 0.4785\varepsilon_{t-4}$$

```{r Q1 - Predictions 2019Q4}
#2019 Q4
pred.2019Q4 = 408.25 + 405.83 - 405.56 + (0.5614 * 0.031927800) - (0.8524 * 0.009630730) - (0.4785 * -0.101088224)
```

Prediction 2020 Q1:$$ y_{t+2} = y_{t+1} + y_{t-2} - y_{t-3} + \varepsilon_{t+2} + 0.5614\varepsilon_{t+1} - 0.8524\varepsilon_{t-2} - 0.4785\varepsilon_{t-3}$$
```{r Q1 - Predictions 2020Q1}
#2020Q1
pred.2020Q1 = pred.2019Q4 + 405.73 - 405.83 - (0.8524 * -0.106133913) - (0.4785 * 0.009630730)
pred.2020Q1
```

Prediction 2020 Q2: $$ y_{t+3} = y_{t+2} + y_{t-1} - y_{t-2} + \varepsilon_{t+3} + 0.5614\varepsilon_{t+2} - 0.8524\varepsilon_{t-1} - 0.4785\varepsilon_{t-2}$$
```{r Q1 - Predictions 2020Q2}
#2020Q3
pred.2020Q2 = pred.2020Q1 + 406.71 - 405.73 - (0.8524 * 0.402908803) - (0.4785 * -0.106133913)
```

Prediction 2020 Q3: $$ y_{t+4} = y_{t+3} + y_{t} - y_{t-1} + \varepsilon_{t+4} + 0.5614\varepsilon_{t+3} - 0.8524\varepsilon_{t} - 0.4785\varepsilon_{t-1}$$
```{r Q1 - Predictions 2020Q3}
#2020Q4
pred.2020Q3 = pred.2020Q2 + 408.25 - 406.71 - (0.8524 * 0.031927800) - (0.4785 * 0.402908803)
```
Prediction Results:
```{r Q1 - Predictions By Hand}
#results
results.df = data.frame(Time=c("2019.4", "2020.1", "2020.2", "2020.3"),
                           Predictions=c(round(pred.2019Q4,2),round(pred.2020Q1,2),round(pred.2020Q2,2),round(pred.2020Q3,2)))
                          
results.df
```

## Question 3 - Executive Summary

The task was to predict the atmospheric concentration of carbon dixiode at Cape Grim, Tasmania, Australia (in parts per million) between 2019 Quarter 4 and 2020 Quarter 3. 

We need to be wary of our predictions and their reliability as we have a time series with only 79 observations. However the model that is used is a good model therefore the predictions should be reasonably reliable.

Several different models were built using observations between 2000 Quarter 1 and 2018 Quarter 3 and used to predict 2018 Quarter 4 to 2019 Quarter 3. Each model's predictions were then compared to the actual values to find the model that produced the most accurate predictions. 
The best predicting model found from this method was then re-run on all the avaliable data and predictions for 2019 Quarter 4 to 2020 Quarter 3 were produced as per the task. 

We predict the carbon dixiode concentration in the atmosphere above Cape Grim in Tasmania, Australia will be:
2019 Quarter 4: 408.60 ppm
2020 Quarter 1: 408.61 ppm
2020 Quarter 2: 409.34 ppm
2020 Quarter 3: 410.76 ppm 
