---
title: "Lab 2: Bike cleaning"
author: "Hasnain Cheena"
date: "9 August 2020"
output: html_document
---

## Cleaning the bike data with refinr

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

The cycle counter data from the first assignment:

```{r}
bike_files<- list.files("../data/BIKES/",pattern=".csv",full=TRUE)
bike_data<-map(bike_files,read_csv)
```

There are some columns that are entirely `NA`: get rid of them with `select_if`

```{r}
notallNA<-function(x) !all(is.na(x))
bike_data <- map(bike_data, select_if, notallNA)
```

Now we could try tidying the data

```{r}
bike_tidy<-map(bike_data, pivot_longer, values_to ="rider_count", names_to ="counter",-Date)
allbikes<-bind_rows(bike_tidy)
allbikes %>% group_by(counter) %>% count() 
```

Even among the first few rows shown it's clear that `Curran St total`, `Curran Street Total`, and 
`Curran Street Total Cyclists` should be the same. 

Load the `refinr` package and read the vignette. Try using `n_gram_merge()` and `key_collision_merge` on the `counter` variable.  Then try the `ignore_strings` argument to the functions, starting with `ignore_strings=c("cycle","counter","total","cyclists")`

```{r include=FALSE}
library(refinr)
sort(unique(n_gram_merge(allbikes$counter,  ignore_strings=c("cycle","counter","total","cyclists", "road", "rd", "street", "st", "drive", "dr", "wst", "west", "sp"))))

sort(unique(key_collision_merge(allbikes$counter, ignore_strings=c("cycle","counter","total","cyclists", "road", "rd", "street", "st", "drive", "dr", "wst", "west", "sp"))))
```
You won't be able to get all the way automatically: there's no way `refinr` can know that "Quay St Vector Arena" and "Quay St Spark Arena" are the same as each other but not the same as "Quay St Totem"

The next step is to write out to a file a two-column table with the distinct original values of `counter` and the merged values. You'll have to edit the second column of this table by hand in a text editor or spreadsheet (it helps if it's in alphabetical order).  Read the table back in and call it something like `name_table`, with column names `counter` and `counter_edited`

```{r Task 1}
#write out table
name_table = data.frame("counter"= sort(unique(allbikes$counter)),
                                "counter_edited" = c(''))
#only use when manually mapping counters
# write_csv(name_table, "name_table.csv")
```

```{r Task 1b}
#read in edited table
name_table = read_csv("name_table.csv")
head(name_table)
```

Now, use a join on `counter` to add the edited name to the `allbikes` data set.   Plot a graph showing the day-of-week pattern for each counter. 

```{r Task 2, fig.width=7, fig.height=100}

#join and add dow variable
full_bikes_df = inner_join(name_table,allbikes, by="counter") %>%
  mutate(dow = substr(Date,start=1, stop=3),
         wday = factor(dow, levels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
         area_location = factor(counter_edited)) 

#sumamrize 
bike_summary_df = full_bikes_df %>%
  filter(!(is.na(wday))) %>%
  group_by(area_location, wday) %>%
  summarise(avg_cyclists = mean(rider_count, na.rm=TRUE)) 

#plot graph
p = ggplot(data=bike_summary_df, aes(x=wday, y=avg_cyclists)) + 
  geom_point() +
  labs(title="Average Number of Cyclists by Day of Week and Location", x="Day of Week", y="Number of Cyclists")

p + facet_grid(rows = vars(area_location))
```

*Note: there's a standalone package called 'OpenRefine' that's better than `refinr` but it's not installed on the lab computers*