---
title: 'Stats 326: Assignment 3'
author: "Hasnain Cheena"
date: "16/04/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(s20x)
```

```{r Data Setup, echo=FALSE}
A4.df = read.table('Ass 4 Data.txt', header=TRUE) 

```

## Question 1

```{r TS1 Explore}
plot.ts(A4.df$TS1, main="TS1")
acf(A4.df$TS1)
pacf(A4.df$TS1)
```

$$ y_t = \rho_{1} y_{t-1} \times \rho_2 y_{t-2} + \varepsilon_t $$

The plot of the series shows clustering indcating positive autocorrelation. The acf shows decay while the pacf shows cut-off at lag 2. This suggests AR(2) is the most suitable model.

```{r TS1 Fit}
TS1.fit = arima(A4.df$TS1, order=c(2,0,0))
TS1.fit

plot(residuals(TS1.fit), main="Residual Series")
acf(residuals(TS1.fit))
```

$$ y_t = 0.5958 y_{t-1} \times 0.2928 y_{t-2} + \varepsilon_t $$ 

```{r TS1 Fit Other Models, echo=FALSE}
TS1.AR3.fit = arima(A4.df$TS1, order=c(3,0,0))
TS1.AR3.fit

TS1.ARMA1.fit = arima(A4.df$TS1, order=c(2,0,1))
TS1.ARMA1.fit
```
The Residual Series appear to be random scatter about 0. The plot of the autocorrelation function of the Residual Series shows no significant lags. 

Other models tried:

AR(3)        AIC = 2857.86    
ARMA(2,10)   AIC = 2889.29    

AR(2) is the best model. This is because all terms in the AR(2) and AR(3) were significant. However, the AIC of the AR(2) is lower than the AR(3) and the AR(2) is a simpler model. 

## Question 2

```{r TS1 Explore}
plot.ts(A4.df$TS2, main="TS2")
acf(A4.df$TS2)
pacf(A4.df$TS2)
```
The plot of the series shows no discrenable pattern. The acf shows cut-off at lag 2 and the pacf shows decay (or persistence). This suggests MA(2) is the most suitable model. 

$$ y_t = \varepsilon _{t} + \alpha_1\varepsilon_{t-1}$$



## Question 3

```{r TS1 Explore}
plot.ts(A4.df$TS3, main="TS3")
acf(A4.df$TS3)
pacf(A4.df$TS3)
```
$$ y_t = \varepsilon _{t}$$

The plot of the series shows no discrenible pattern. The acf and pacf show no significant lags. This suggests the series is White Noise. 

```{r TS3 Fit}
TS3.fit = arima(A4.df$TS3, order=c(0,0,0))
TS3.fit

plot(residuals(TS3.fit), main="Residual Series")
acf(residuals(TS3.fit))
```
$$ y_t = \varepsilon _{t}$$

```{r TS3 Fit Other Models, echo=FALSE}
TS3.AR1.fit = arima(A4.df$TS3, order=c(1,0,0))
TS3.AR1.fit

TS3.MA1.fit = arima(A4.df$TS3, order=c(0,0,1))
TS3.MA1.fit

```

The Residual Series appears to be random scatter about 0. The plot of the autocorrelation function of the Residual Series shows no significant lags. 

Other models tried: 
AR(1)    AIC: 2839.17
MA(1)    AIC: 2839.14 


## Question 4

```{r TS4 Explore}
plot.ts(A4.df$TS4, main="TS4")
acf(A4.df$TS4)
pacf(A4.df$TS4)
```
$$ y_t = \rho_1 y_{t-1} + \alpha_1 \varepsilon _{t-1} + \varepsilon _{t}$$

The plot of the series shows clustering indicating positive autocorrelation. Both the acf and pacf show decay. This suggests an ARMA(p,q) is an appropriate model. However from the plots we have no indication of what order of the ARMA model and thus I began ARMA(1,1).

```{r TS4 Fit}
TS4.fit = arima(A4.df$TS4, order=c(1,0,1))
TS4.fit

plot(residuals(TS4.fit), main="Residual Series")
acf(residuals(TS4.fit))
```

$$ y_t = 0.8974 y_{t-1} +  0.9121 \varepsilon _{t-1} + \varepsilon _{t}$$

```{r TS4 Fit Other Models, echo=FALSE}
TS4.AR2MA1.fit = arima(A4.df$TS4, order=c(2,0,1))
TS4.AR2MA1.fit

TS4.AR1MA2.fit = arima(A4.df$TS4, order=c(1,0,2))
TS4.AR1MA2.fit
```

The Residual Series appears to be random scatter about 0. The plot of the autocorrelation function of the Residual Series shows 2 weakly signficant lags at lags 20 and 21. As they are weakly significant, they are not a concern. 

Other models tried: 
ARMA(2,1)    AIC: 2833.86 
ARMA(1,2)    AIC: 2833.73

ARMA(1,1) is the best model. Even though all terms are significant in ARMA(2,1) and ARMA(1,2), ARMA(1,1) has a lower AIC score and is a simpler model. 


## Question 5

```{r TS5 Explore}
plot.ts(A4.df$TS5, main="TS5")
acf(A4.df$TS5)
pacf(A4.df$TS5)
```
$$ y_t = \rho_1 y_{t-1} + \alpha_1 \varepsilon _{t-1} + \varepsilon _{t}$$

The plot of the series shows clustering indicating positive autocorrelation. Both the acf and pacf show decay/persistance. This suggests an ARMA(p,q) is an appropriate model. However from the plots we have no indication of what order of the ARMA model and thus I began ARMA(1,1).

```{r TS5 Fit}
TS5.fit = arima(A4.df$TS5, order=c(1,0,1))
TS5.fit

plot(residuals(TS5.fit), main="Residual Series")
acf(residuals(TS5.fit))
```

$$ y_t = 0.9674 y_{t-1} +  0.1876 \varepsilon _{t-1} + \varepsilon _{t}$$
The Residual Series appears to be random scatter about 0. The plot of the autocorrelation function of the Residual Series shows 4 significant lags at lags 2, 3, 9 and 15. 

Better Model: 
```{r TS5 Fit Better Model}
TS5.AR5MA1.fit = arima(A4.df$TS5, order=c(5,0,1))
TS5.AR5MA1.fit
```

$$ y_t = 1.6551 y_{t-1}  -0.6328 y_{t-2} - 0.2822 y_{t-3} + 0.4033 y_{t-4} - 0.1634 y_{t-5} - 0.4363 \varepsilon _{t-1} + \varepsilon _{t}$$

```{r TS4 Fit Better Model Plots}
plot(residuals(TS5.AR5MA1.fit), main="Residual Series")
acf(residuals(TS5.AR5MA1.fit))

```
The Residual Series appears to be random scatter about 0. The plot of the autocorrelation function of the residual series shows no signficant lags. 

Therefore, ARMA(5,1) was decided as the best model for this series. 


