---
title: 'Stat 369 Lab 6: Autism trees'
date: "24 September 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rpart)
```

This lab looks at data collected to try to detect autism spectrum disorder (ASD) biochemically. We use trees to detect ASD.

Submit a knitted html or pdf of your answer to Canvas.

## Tasks

#### Read in Data
```{r Read Data}
vnames <- read_csv("S1Dataset.csv",n_max=0)
autism<- read_csv("S1Dataset.csv",skip=2)
names(autism)<-names(vnames)
```

#### Task 1 
1. Use `rpart()` to grow a tree that distinguishes the ASD and NEU groups, and give the confusion matrix and the proportion of people (from these two groups) correctly classified. Plot the tree (showing labels -- make sure you set the margin in the plot call to make room for the labels).

```{r Task 1}
autism.tree = rpart(Group~., data=autism)
autism.tree
```


2. The dataset makes the classifier look good, because the number of ASD and neurotypical participants are equal.  In reality, ASD affects about 1.5% of people.  Run `rpart()` with the prior argument (see the help page), to build a tree based on population prior probabilities of 0.015 and 0.985. Report the confusion matrix and the proportion of ASD and NEU groups correctly classified. Plot the tree.

```{r setup, Read Data}

```

3. Suppose that false negative classifications (missing ASD) are thought to be more important than false positive (suspecting ASD). Run `rpart()` with the `prior` argument as in the previous question and also with the `loss` argument saying that false negatives are 10 times as bad as false positives.
Report the confusion matrix and the proportion of ASD and NEU  groups correctly classified. Plot the tree. Did you get the result you expected? 


4. Use this final tree to predict for the SIB group (all of whom are actually neurotypical). What proportion are correctly classified?
